Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces

### 2021-03-26 analysis/01_clean-data.Rmd
Convert everything to targets

```{r load-libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(countrycode)
library(here)
```

# Load raw data

```{r load-raw-data, cache=TRUE, warning=FALSE, message=FALSE}
# V-Dem pandemic backsliding
pandem_raw <- read_csv(here("data", "raw_data", "pandem", "csv_files", "PanDem_ts_V5.csv"))

# ICCPR derogations
derogations_raw <- read_csv(here("data", "raw_data", "derogations", "Pandemic backsliding - Sheet1.csv"))

# V-Dem
vdem_raw <- read_rds(here("data", "raw_data", "Country_Year_V-Dem_Full+others_R_v10",
                         "V-Dem-CY-Full+Others-v10.rds")) %>% as_tibble()
```

# Clean up data

```{r clean-data}
emlaw_lookup <- tribble(
  ~emlaw, ~emlaw_fct, ~emlaw_cond,
  1, "Declaration in existing legal framework", "Emergency response with legal instruments",
  2, "Declaration in existing legal framework, distinguishes public health emergency", "Emergency response with legal instruments",
  3, "Declaration of disaster where different from emergency", "Emergency response with legal instruments",
  4, "Declaration using other legislation", "Emergency response with legal instruments",
  5, "Emergency response without legal instruments", "Emergency response without legal instruments",
  6, "No emergency response", "No emergency response",
  7, "Other", "Other"
)

regime_lookup <- tribble(
  ~v2x_regime, ~regime,
  0, "Closed autocracy",
  1, "Electoral autocracy",
  2, "Electoral democracy",
  3, "Liberal democracy"
) %>% 
  mutate(regime = fct_inorder(regime, ordered = TRUE))

type_lookup <- tribble(
  ~type, ~type_clean,
  "A", "Derogated as intended",
  "B", "Derogated; measures not temporary, proportional, or necessary",
  "C", "Emergency declared; no formal derogation",
  "D", "No emergency; no formal derogation",
  "E", "Has not ratified ICCPR"
) %>% 
  mutate(type_clean_letter = paste0(type, ": ", type_clean)) %>% 
  mutate(across(starts_with("type"), ~fct_inorder(., ordered = TRUE)))

pandem_clean <- pandem_raw %>% 
  mutate(ndrights_fct = factor(ndrights, levels = 0:1, 
                               labels = c("No violations", "Violations of non-derogable rights"))) %>% 
  group_by(country_name) %>% 
  mutate(emlaw_min = min(emlaw),
         pandem_max = max(pandem),
         panback_max = max(panback)) %>% 
  ungroup() %>% 
  left_join(emlaw_lookup, by = "emlaw") %>% 
  left_join(rename_with(emlaw_lookup, ~str_c(., "_min"), everything()), by = "emlaw_min") %>% 
  left_join(regime_lookup, by = "v2x_regime") %>% 
  mutate(emlaw_fct = factor(emlaw_fct, levels = emlaw_lookup$emlaw_fct, ordered = TRUE),
         emlaw_cond = factor(emlaw_cond, levels = unique(emlaw_lookup$emlaw_cond), ordered = TRUE)) %>% 
  mutate(emlaw_min_fct = factor(emlaw_fct, levels = emlaw_lookup$emlaw_fct, ordered = TRUE),
         emlaw_min_cond = factor(emlaw_cond, levels = unique(emlaw_lookup$emlaw_cond), ordered = TRUE)) %>% 
  mutate(cowcode = countrycode(country_name, "country.name", "cown",
                               custom_match = c("Serbia" = 345,
                                                "Hong Kong" = 997)),
         iso3 = countrycode(country_name, "country.name", "iso3c"))

pandem_single <- pandem_clean %>% 
  filter(Quarter == "Q4")
glimpse(pandem_single)

derogations_clean <- derogations_raw %>% 
  rename(ratified = `Did they ratify/are they a member? (1=Yes, 0=No, 99= Not possible)`,
         derogate = `Did they officially derogate?`,
         derogation_start = `Derogation start date`,
         derogation_end = `Derogation end date`,
         country_name = Country) %>% 
  mutate(country_name = recode(country_name, "Columbia" = "Colombia")) %>% 
  mutate(cowcode = countrycode(country_name, "country.name", "cown",
                               custom_match = c("Serbia" = 345,
                                                "State of Palestine" = 998,
                                                "Cook Islands" = NA,
                                                "Niue" = NA)),
         iso3 = countrycode(country_name, "country.name", "iso3c"))
glimpse(derogations_clean)

vdem_clean <- vdem_raw %>% 
  filter(year > 2015) %>% 
  rename(cowcode = COWcode) %>% 
  mutate(cowcode = case_when(
    country_name == "Hong Kong" ~ 997,
    country_name == "Palestine/West Bank" ~ 998,
    TRUE ~ cowcode
  )) %>% 
  mutate(iso3c = countrycode(country_name, "country.name", "iso3c",
                             custom_match = c("Kosovo" = "XKK",
                                              "Somaliland" = "XSO",
                                              "Zanzibar" = "EAZ"))) %>% 
  select(country_name, year, cowcode, iso3c, 
         # Civil society stuff
         v2cseeorgs,  # CSO entry and exit
         v2csreprss,  # CSO repression
         v2cscnsult,  # CSO consultation
         v2csprtcpt,  # CSO participatory environment
         v2csgender,  # CSO women's participation
         v2csantimv,  # CSO anti-system movements
         v2xcs_ccsi,  # Core civil society index (entry/exit, repression, participatory env)
         # Human rights and politics
         # Political corruption index (less to more, 0-1) (public sector +
         # executive + legislative + judicial corruption)
         v2x_corr,
         # Rule of law index
         v2x_rule,
         # Rights indexes
         v2x_civlib,  # Civil liberties index
         v2x_clphy,  # Physical violence index
         v2x_clpriv,  # Private civil liberties index
         v2x_clpol,  # Political civil liberties index
         # Democracy
         v2x_polyarchy
  )
glimpse(vdem_clean)

pandem_derog <- derogations_clean %>% 
  left_join(pandem_single, by = c("country_name", "cowcode", "iso3")) %>% 
  left_join(filter(vdem_clean, year == 2019), by = c("country_name", "cowcode")) %>% 
  filter(!is.na(X1)) %>%
  mutate(has_derogation_end = case_when(
    is.na(derogation_end) ~ FALSE,
    str_detect(derogation_end, "N/A") ~ FALSE,
    TRUE ~ TRUE
  )) %>% 
  mutate(has_emergency_end = emlimit == 1) %>% 
  mutate(nonproportional = Type1 > 1 | Type2 == 1 | Type3 > 1) %>% 
  mutate(type = case_when(
    derogate == 1 & has_derogation_end & !nonproportional ~ "A",
    derogate == 1 & (!has_derogation_end | nonproportional) ~ "B",
    derogate == 0 & ratified == 1 & emlaw < 5 ~ "C",
    derogate == 0 & ratified == 1 & emlaw >= 5 ~ "D",
    TRUE ~ "E"
  )) %>% 
  left_join(type_lookup, by = "type")
glimpse(pandem_derog)
```

# Save data for later use

```{r save-data}
saveRDS(pandem_clean, here("data", "derived_data", "pandem_clean.rds"))
saveRDS(pandem_single, here("data", "derived_data", "pandem_single.rds"))
saveRDS(derogations_clean, here("data", "derived_data", "derogations_clean.rds"))
saveRDS(vdem_clean, here("data", "derived_data", "vdem_clean.rds"))
saveRDS(pandem_derog, here("data", "derived_data", "pandem_derog.rds"))
```

>>>>

# `targets` pipeline

Here's the general process for building and running this analysis. This is all done with [the magical **`targets`** package](https://docs.ropensci.org/targets/), which orchestrates all the dependencies automatically.

```{r show-targets-pipeline, echo=FALSE}
withr::with_dir(here::here(), {
  targets::tar_glimpse()
})
```

# Data cleaning

All the data processing is handled with dataset-specific functions that live in `R/funs_data-cleaning.R`, which **`targets`** then runs as needed. For the sake of transparency, here's that code:

```{r, code=xfun::read_utf8(here::here("R", "funs_data-cleaning.R")), eval=FALSE}
```

### 2021-03-26 analysis/02_descriptive-analysis.Rmd
Convert everything to targets

```{r load-libraries-data, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(ggbeeswarm)
library(patchwork)
library(here)

source(here("lib", "graphics.R"))

@@

pandem_clean <- readRDS(here("data", "derived_data", "pandem_clean.rds"))
pandem_single <- readRDS(here("data", "derived_data", "pandem_single.rds"))
derogations_clean <- readRDS(here("data", "derived_data", "derogations_clean.rds"))
vdem_clean <- readRDS(here("data", "derived_data", "vdem_clean.rds"))
pandem_derog <- readRDS(here("data", "derived_data", "pandem_derog.rds"))
```

>>>>

```{r load-libraries-data, message=FALSE, warning=FALSE}
library(tidyverse)
library(targets)
library(scales)
library(ggbeeswarm)
library(patchwork)
library(here)

@@

# Load data
# Need to use this withr thing because tar_read() and tar_load() need to see the
# _targets folder in the current directory, but this .Rmd file is in a subfolder
withr::with_dir(here::here(), {
  source(tar_read(plot_funs))

  tar_load(pandem_clean)
  tar_load(pandem_single)
  tar_load(derogations_clean)
  tar_load(vdem_clean)
  tar_load(pandem_derog)
})
```

### 2021-03-26 manuscript/manuscript.Rmd
Convert everything to targets

---
title: "Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces"
short-title: "Derogations and Democratic Backsliding"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: March 21, 2021
published: This is an early draft of a working paper.
code-repo: Access the code and data at <https://github.com/andrewheiss/covid-derogations>.
abstract: >-
  In an effort to combat the COVID-19 pandemic, most governments have imposed restrictions on civic freedoms in the interest of public health. In many cases, these emergency measures have been temporary and proportional in terms of controling the pandemic. In many other instances, however, governments have used these restrictions to suppress opposition and more permanently restrict civic space. Systematically measuring the consequences of COVID restrictions, however, is a difficult task. We examine two possible quantitative measures of the relationship between of COVID restrictions and civil society space. First, we use the Variety of Democracy project's newly released Pandemic Violations of Democratic Standards Index to explore if and how civil society restrictions predict pandemic backsliding. Second, while many countries sign international human rights treaties that ostensibly bind states to respect rights, several treaties allow for emergency derogations from these obligations. We tabulate formal human rights treaty derogations due to the pandemic and explore whether these emergency measures led to lasting declines in associational and human rights. We hope that our exploration of these two measures will provide rich data-based descriptions of the relationship between COVID restrictions and civic space that will allow for more causal work in the future.
keywords: "international human rights treaties, civil society, COVID-19, democracy"
reference-section-title: References
bibliography: bibliography.bib
csl: pandoc/csl/chicago-author-date.csl
title-page: true
endnotes: false
endfloat: false
mainfont: Spectral
sansfont: Noto Sans
editor_options: 
  chunk_output_type: console
---

@@

# Load libraries
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
library(janitor)
library(ggbeeswarm)
library(patchwork)
library(sf)
library(here)

source(here("lib", "graphics.R"))

@@

# Load data
pandem_clean <- readRDS(here("data", "derived_data", "pandem_clean.rds"))
pandem_single <- readRDS(here("data", "derived_data", "pandem_single.rds"))
derogations_clean <- readRDS(here("data", "derived_data", "derogations_clean.rds"))
vdem_clean <- readRDS(here("data", "derived_data", "vdem_clean.rds"))
pandem_derog <- readRDS(here("data", "derived_data", "pandem_derog.rds"))
civicus_clean <- readRDS(here("data", "derived_data", "civicus_clean.rds"))

world_map <- read_sf(here("data", "raw_data", "ne_110m_admin_0_countries",
                          "ne_110m_admin_0_countries.shp")) %>% 
  filter(ISO_A3 != "ATA")

@@

```{r civicus-map, fig.cap="2020 CIVICUS Monitor civic space ratings"}
map_with_civicus <- world_map %>% 
  # Fix some Natural Earth ISO weirdness
  mutate(ISO_A3 = ifelse(ISO_A3 == "-99", as.character(ISO_A3_EH), as.character(ISO_A3))) %>% 
  mutate(ISO_A3 = case_when(
    .$ISO_A3 == "GRL" ~ "DNK",
    .$NAME == "Norway" ~ "NOR",
    .$NAME == "Kosovo" ~ "XKK",
    TRUE ~ ISO_A3
  )) %>% 
  left_join(civicus_clean, by = c("ISO_A3" = "iso3c"))

ggplot() +
  geom_sf(data = map_with_civicus, aes(fill = fct_rev(category)), size = 0.15, color = "black") +
  coord_sf(crs = st_crs("ESRI:54030")) +  # Robinson
  scale_fill_viridis_d(option = "plasma", end = 0.9, 
                       na.translate = FALSE, name = "Civic space") +
  theme_pandem_map() + 
  theme(legend.key.size = unit(0.7, "lines"))
```

@@

if (!is_docx) {
  tbl_case_summary %>% 
    kbl(align = c("l", "l", "c", "c", "c", "l"),
        caption = tbl_case_summary_caption,
        booktabs = TRUE, linesep = "\\addlinespace") %>% 
    kable_styling() %>% 
    column_spec(1, width = "0.1\\\\textwidth") %>% 
    column_spec(2, width = "0.12\\\\textwidth") %>% 
    column_spec(3, width = "0.09\\\\textwidth") %>% 
    column_spec(4, width = "0.11\\\\textwidth") %>% 
    column_spec(5, width = "0.09\\\\textwidth") %>% 
    column_spec(6, width = "0.35\\\\textwidth") %>% 
    footnote(general = tbl_case_notes)
} else {
  tbl_case_summary %>% 
    hux() %>% 
    set_font_size(10) %>%
    set_all_padding(3) %>% 
    set_align(everywhere, 3:5, "center") %>% 
    set_align(everywhere, c(1, 2, 6), "left") %>% 
    set_bold(1, everywhere) %>%
    set_bottom_border(1, everywhere) %>%
    huxtable::add_footnote(paste(tbl_case_notes, collapse = " • ")) %>% 
    huxtable::set_caption(tbl_case_summary_caption) %>% 
    huxtable::as_flextable()
}
```

>>>>

---
title: "Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces"
short-title: "Derogations and Democratic Backsliding"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: March 21, 2021
published: This is an early draft of a working paper.
code-repo: Access the code and data at <https://github.com/andrewheiss/covid-derogations>.
abstract: >-
  In an effort to combat the COVID-19 pandemic, most governments have imposed restrictions on civic freedoms in the interest of public health. In many cases, these emergency measures have been temporary and proportional in terms of controling the pandemic. In many other instances, however, governments have used these restrictions to suppress opposition and more permanently restrict civic space. Systematically measuring the consequences of COVID restrictions, however, is a difficult task. We examine two possible quantitative measures of the relationship between of COVID restrictions and civil society space. First, we use the Variety of Democracy project's newly released Pandemic Violations of Democratic Standards Index to explore if and how civil society restrictions predict pandemic backsliding. Second, while many countries sign international human rights treaties that ostensibly bind states to respect rights, several treaties allow for emergency derogations from these obligations. We tabulate formal human rights treaty derogations due to the pandemic and explore whether these emergency measures led to lasting declines in associational and human rights. We hope that our exploration of these two measures will provide rich data-based descriptions of the relationship between COVID restrictions and civic space that will allow for more causal work in the future.
keywords: "international human rights treaties, civil society, COVID-19, democracy"
reference-section-title: References
bibliography: bibliography.bib
csl: pandoc/csl/chicago-author-date.csl
title-page: true
endnotes: false
endfloat: false
mainfont: Cochineal
sansfont: Noto Sans
editor_options: 
  chunk_output_type: console
---

@@

# Load libraries
library(tidyverse)
library(targets)
library(kableExtra)
library(scales)
library(janitor)
library(ggbeeswarm)
library(patchwork)
library(sf)
library(here)

# Load data
# Need to use this withr thing because tar_read() and tar_load() need to see the
# _targets folder in the current directory, but this .Rmd file is in a subfolder
withr::with_dir(here::here(), {
  source(tar_read(plot_funs))

  tar_load(civicus_map_data)
  tar_load(pandem_derog)
  tar_load(pandem_single)
  tar_load(derogations_clean)
  tar_load(civicus_clean)
})

@@

@@

```{r civicus-map, fig.cap="2020 CIVICUS Monitor civic space ratings"}
ggplot() +
  geom_sf(data = civicus_map_data, aes(fill = fct_rev(category)), size = 0.15, color = "black") +
  coord_sf(crs = st_crs("ESRI:54030")) +  # Robinson
  scale_fill_viridis_d(option = "plasma", end = 0.9, 
                       na.translate = FALSE, name = "Civic space") +
  theme_pandem_map() + 
  theme(legend.key.size = unit(0.7, "lines"))
```

@@

if (!is_docx) {
  tbl_case_summary %>% 
    kbl(align = c("l", "l", "c", "c", "c", "l"),
        caption = tbl_case_summary_caption,
        booktabs = TRUE, linesep = "\\addlinespace") %>% 
    kable_styling(font_size = 8.5) %>% 
    row_spec(0, bold = TRUE) %>% 
    column_spec(1, width = "0.1\\\\textwidth") %>% 
    column_spec(2, width = "0.12\\\\textwidth") %>% 
    column_spec(3, width = "0.09\\\\textwidth") %>% 
    column_spec(4, width = "0.11\\\\textwidth") %>% 
    column_spec(5, width = "0.09\\\\textwidth") %>% 
    column_spec(6, width = "0.35\\\\textwidth") %>% 
    footnote(general = tbl_case_notes)
} else {
  tbl_case_summary %>% 
    hux() %>% 
    set_font_size(10) %>%
    set_all_padding(3) %>% 
    set_align(everywhere, 3:5, "center") %>% 
    set_align(everywhere, c(1, 2, 6), "left") %>% 
    set_bold(1, everywhere) %>%
    set_bottom_border(1, everywhere) %>%
    huxtable::add_footnote(paste(tbl_case_notes, collapse = " • ")) %>% 
    huxtable::set_caption(tbl_case_summary_caption) %>% 
    huxtable::as_flextable()
}
```
